---
title: "CSCI 4148 - Practical 0"
author: "Joel Richards"
output:
  html_document:
    df_print: paged
  html_notebook:
    fig_width: 8
    fig_height: 12
    highlight: kate
    theme: paper
---

```{r}
library(tidyverse) 
library(datasauRus)
```

## Data

### Q1

**Based on the help file, how many rows and how many columns does the datasaurus_dozen file have? What are the variables included in the data frame? (this can be hardcoded)**

A: the data frame has 1846 rows and 3 variables.

## Data Visualization and Summary

```{r}
datasaurus_dozen %>%
  count(dataset)
```

***

### Q2

**Plot y vs. x for the dino dataset. Then, calculate the correlation coefficient between x and y for this dataset.**

```{r}
dino_data <- datasaurus_dozen %>%
  filter(dataset == "dino")

ggplot(data = dino_data, mapping = aes(x=x, y=y)) + geom_point()
```

***

### Q3

**Plot y vs. x for the star dataset. You can (and should) reuse code we introduced above, just replace the dataset name with the desired dataset. Then, calculate the correlation coefficient between x and y for this dataset. How does this value compare to the r of dino?**

```{r}
star_data <- datasaurus_dozen %>%
  filter(dataset == "star")

ggplot(data = star_data, mapping = aes(x = x, y = y)) + 
  geom_point()

dino_data %>%
  summarise(r = cor(x, y))

star_data %>%
  summarise(r = cor(x, y))
```

A: The correlation coefficients of the dino and star datasets are similar, at -0.064 and -0.063, respetively. Both indicate a very poor linear relationship.

***

### Q4

**Plot y vs. x for the circle dataset. You can (and should) reuse code we introduced above, just replace the dataset name with the desired dataset. Then, calculate the correlation coefficient between x and y for this dataset. How does this value compare to the r of dino?**

```{r}
star_data <- datasaurus_dozen %>%
  filter(dataset == "circle")

ggplot(data = star_data, mapping = aes(x = x, y = y)) + 
  geom_point()

star_data %>%
  summarise(r = cor(x, y))
```

A: Again, the correlation coefficient is very low. Still on the order of 0.06.

***

### Q5

**Finally, let’s plot all datasets at once. In order to do this we will make use of facetting.**

```{r fig.height=8, fig.width=8}
ggplot(datasaurus_dozen, aes(x=x, y=y), color=dataset) +
  geom_point() + 
  facet_wrap(~ dataset, ncol=3) +
  theme(legend.position="none")

datasaurus_dozen %>%
  group_by(dataset) %>%
  summarise(r=cor(x, y))
  
```


